---
title: "Attendance & GPA Analysis"
author: "Tasnim Khan"
date: "2026-06-16"
output: html_document
---

Loading Libraries
```{r}
library(dplyr)
library(tidyverse)
```

Loading files
```{r}
#Set your working directory
roster = read.csv("File1.csv")
grade_8_outcomes = read.csv("File2.csv")
hs_outcomes = read.csv("File3.csv")
```

```{r}

#Adding GPA column to Roster table
roster_with_gpa <- roster %>%
  left_join(grade_8_outcomes, by = "studID")

```


Question 3: The following questions pertain to the group of students that our Network wants to focus on supporting, specifically students with a high attendance rate (above 90%) but low GPAs (below 3.0). Below is a function that could answer these questions for any combination of attendance rate and GPA category. Then it's been ran for the group of students with high attendance and low GPA.
```{r}

#Cleaning for next steps
master <- roster_with_gpa %>%
  filter(!is.na(GPA) & GPA != 8.0)


# Add a new column that calculates the attendance rate, format for 2 decimal places
master$attend_rate <- (master$daysAttend / master$daysEnroll) * 100
master$attend_rate <- round((master$daysAttend / master$daysEnroll) * 100, 2)

#Creating categorical variables
master_cats <- master %>% 
  mutate(
    attend_cat = case_when(
      attend_rate < 90  ~ "Low",
      attend_rate >= 90 ~ "High",
      TRUE ~ NA_character_ # Handles any missing/NA rows
    ),
      GPA_cat = case_when(
      GPA < 3.00000  ~ "Low",
      GPA >= 3.00000 ~ "High",
    ) 
  )
master_cats_hs <- master_cats %>%
      left_join(hs_outcomes, by = "studID")

#For the function we define the following assumptions:
#Remove NAs and 8.0s in the GPA column
#Make a attendance rate percentage column
#Has high school outcomes with hsID and gradStatus columns
#Created a binary categorical variable for attendance rate and GPA
  #High attendance is 90 and above, Low attendance is below 90
  #High GPA is 3.0 and above, Low GPA is below 3.0

```

Writing functions
```{r}
#3a: Has the percentage of students in this group changed over time (across different cohorts of 8th graders)?
group_over_time <- function(data, attend_binary, GPA_binary) {
  data %>%
    group_by(schoolYear8th) %>%
    summarize(
      total_students = n(),
      group_students = sum(attend_cat == attend_binary & GPA_cat == GPA_binary),
      percent_groups = round(group_students / total_students * 100, 2)
    )
  
}

specific_results <- group_over_time(
  data = master_cats_hs, 
  attend_binary = "High",
  GPA_binary = "Low"
)

print(specific_results)

```

```{r}
#3b: What is the race/ethnicity composition of the students in this group, and how does it compare to the composition of the entire population of students?

race_composition <- function(data, attend_binary, GPA_binary) {
  
  selected_group <- data %>%
    filter(attend_cat == attend_binary & GPA_cat == GPA_binary) %>%
    count(raceEth) %>%
    mutate(
      group = "Selected Group",
      percent = round(n / sum(n) * 100, 2)
    )
  
  all_students <- data %>%
    count(raceEth) %>%
    mutate(
      group = "All Students",
      percent = round(n / sum(n) * 100, 2)
    )
  
  bind_rows(selected_group, all_students)
}

specific_results <- race_composition(
  data = master_cats_hs, 
  attend_binary = "High",
  GPA_binary = "Low"
  
)

print(specific_results)
  
```

```{r}
#3c: What is the high school graduation rate of students who were in this group in eighth grade? 

graduation_rate <- function(data, attend_binary, GPA_binary) {
  
  data %>%
    filter(attend_cat == attend_binary & GPA_cat == GPA_binary) %>%
    filter(!is.na(gradStatus)) %>%
    summarise(
      total_students = n(),
      graduates = sum(gradStatus == 1),
      graduation_rate = round(mean(gradStatus == 1) * 100, 2)
    )
}

specific_results <- graduation_rate(
  data = master_cats_hs, 
  attend_binary = "High",
  GPA_binary = "Low"
  
)

print(specific_results)


```

```{r}
#3d: How does the high school graduation rate of students in this category vary across middle schools? 

graduation_by_school <- function(data, attend_binary, GPA_binary) {
  
  data %>%
    filter(attend_cat == attend_binary & GPA_cat == GPA_binary) %>%
    filter(!is.na(gradStatus)) %>%
    group_by(schoolID) %>%
    summarise(
      total_students = n(),
      graduates = sum(gradStatus == 1, na.rm = TRUE),
      graduation_rate = round(mean(gradStatus == 1, na.rm = TRUE) * 100, 2)
    ) %>%
    arrange(desc(graduation_rate))
}

specific_results <- graduation_by_school(
  data = master_cats_hs, 
  attend_binary = "High",
  GPA_binary = "Low"
  
)

print(specific_results)

```

```{r}
#How does it vary across cohorts (years)?

graduation_by_cohort <- function(data, attend_binary, GPA_binary) {
  
  data %>%
    filter(attend_cat == attend_binary & GPA_cat == GPA_binary) %>%
    filter(!is.na(gradStatus)) %>%
    group_by(schoolYear8th) %>%
    summarise(
      total_students = n(),
      graduates = sum(gradStatus == 1, na.rm = TRUE),
      graduation_rate = round(mean(gradStatus == 1, na.rm = TRUE) * 100, 2)
    )
}

specific_results <- graduation_by_cohort(
  data = master_cats_hs, 
  attend_binary = "High",
  GPA_binary = "Low"
  
)

print(specific_results)


```

Fuse all the mini functions into one primary function
```{r}
analyze_group <- function(data, attend_binary, GPA_binary) {
  list(
    group_over_time = group_over_time(data, attend_binary, GPA_binary), 
    race_composition = race_composition(data, attend_binary, GPA_binary), 
    graduation_rate = graduation_rate(data, attend_binary, GPA_binary), 
    graduation_by_school = graduation_by_school(data, attend_binary, GPA_binary), 
    graduation_by_cohort = graduation_by_cohort(data, attend_binary, GPA_binary)
  )
}

specific_results <- analyze_group(
  data = master_cats_hs, 
  attend_binary = "High",
  GPA_binary = "Low"
  
)

print(specific_results)

```


Clean output
```{r}
analyze_group <- function(data, attend_binary, gpa_binary) {
  
  cat("\n==============================\n")
  cat("Analysis for:", attend_binary, "Attendance +", gpa_binary, "GPA\n")
  cat("==============================\n\n")
  
  cat("1. Percentage of students in this group over time\n")
  print(group_over_time(data, attend_binary, gpa_binary))
  
  cat("\n2. Race/ethnicity composition\n")
  print(race_composition(data, attend_binary, gpa_binary))
  
  cat("\n3. Overall high school graduation rate\n")
  print(graduation_rate(data, attend_binary, gpa_binary))
  
  cat("\n4. Graduation rate by middle school\n")
  print(graduation_by_school(data, attend_binary, gpa_binary))
  
  cat("\n5. Graduation rate by cohort\n")
  print(graduation_by_cohort(data, attend_binary, gpa_binary))
}

analyze_group(
  data = master_cats_hs,
  attend_binary = "High",
  gpa_binary = "Low"
)


```
